---
title: "Title Goes Here"
author:
  - author 1
  - author 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: journal-of-archaeological-science.csl
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---


<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(modelextinctionideas)
```

# Introduction

Here is a citation [@Marwick2017]

# Background

```{r load-data}

# reload --------------------------------------------------------------------
# this is the fast way to resume:
items_df <- readRDS("../data/derived_data/saa2018-data-df.rds")

library(tidyverse)
items_df <-
items_df %>%
  # remove branding
  filter(authors != "FN Clarivate Analytics Web of Science\nVR 1.0") %>%
  # uniques only
  group_by(authors, title, journal) %>%
  filter(row_number() == 1)

items_wth_refs <-
  items_df %>%
  filter(!is.na(refs)) %>%
  filter(!is.na(year))

# how many items do we have?
hm_items <- nrow(items_wth_refs) # 12133

# span what time period?
time_range <- range(items_wth_refs$year) # 1905 2017

```

We have `r hm_items` articles in our study. The publication dates range from `r time_range[1]` to  `r time_range[2]`.

The distribution of articles is strongly skewed to more recent years, with a rapid increase after the mid-2000s.

```{r}
items_wth_refs %>% 
  group_by(year) %>% 
  tally() %>% 
ggplot(aes(year,
           n)) +
  geom_col() +
  theme_bw() +
  ggtitle("Archaeology articles per year")
```

Let's look at the citations in each of these articles. As expected, the number of citations per year closely follows the distribution of articles per year.

On the log-scaled plot we can see that we have several gaps in our data before 1975. Why is that?

```{r}

# df of refs with year that each ref appears, and ID of citing article
library(stringr)
library(purrrlyr)

# get data frame with refs for each article
# each row is a single citation,
# with info of article that has that citation
xx <-
items_wth_refs %>%
  by_row(..f = function(this_row){
   unlist(str_split(this_row$refs, "\n"))
  }) %>%
  unnest(.out) %>%
  # remove non-refs
  filter(!grepl("^[A-Z0-9]{2} .*|^ER$", .out)) %>%
  filter( .out != "")

# how many citations per year?
xxx <-
xx %>%
  group_by(year) %>%
  tally()

# plot, why do we have some gaps? in the 1970s?
ggplot(xxx,
       aes(year,
            n)) +
  geom_col() +
  #scale_y_log10() +
  theme_bw() +
  ggtitle("Citations per year in archaeology articles")

library(scales)
ggplot(xxx,
       aes(year,
            n)) +
  geom_col() +
  scale_y_log10(breaks = trans_breaks('log10', function(x) 10^x),
                  labels = trans_format('log10', math_format(10^.x))) +
  theme_bw() +
  ggtitle("Citations per year in archaeology articles (log scale)")
```

We can zoom in on one article, Binford's "Archaeology as Anthropology", published in 1962. The plot shows the proportion of all citations in each year that are Binford 1962. 

At a glance, we have three phases, roughly, high-frequency citations during 1962-1980, medium frequency during 1980-1990, and low-frequency 1990-present.

```{r}

# age distribution of "Archaeology as Anthropology" citations, as a proportion
# of all citations in each year
binford_1962_cites_per_year <-
  xx %>%
  filter(grepl("BINFORD LR, 1962, AM ANTIQUITY, V28, P217,", .out)) %>%
    group_by(year) %>%
    summarise(count = n()) %>%
    left_join(xxx, by = "year") %>%
    mutate(prop = count / n) %>%
    distinct(year, prop, n)

# plot, why no citations throughout the 1970s?
binford_1962_cites_per_year %>%
  ggplot(aes(year, prop)) +
  geom_col() +
  theme_bw() +
  ylab("Proportion of all citations per year") +
  ggtitle("Citations per year of Binford 1962")
```

# Methods

# Results

# Discussion

# Conclusion

# Acknowledgements

# References 

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```
